untitled

### 2016-05-06 appendix.tex
updated notes

Consider axisymmetric disturbances of the form
\begin{align}
  \delta\rho(r,z)\exp{\left(-\ii\sigma\right)}. 
\end{align}
Linearized equations for dusty gas with $\tstop=0$, and using the
pressure equation in place of the dust-fraction
(Eq. \ref{masseq}---\ref{momeq}, Eq. \ref{eff_energy}):

\begin{align}
  \ii\sigma\frac{\delta\rho}{\rho} &= \nabla\cdot\dd\bm{v} +
  \dd\bm{v}\cdot\nabla\ln{\rho},\\
  \ii\sigma\frac{\delta P}{P} &= \nabla\cdot\dd\bm{v} +
  \dd\bm{v}\cdot\nabla\ln{P} - \dd\bm{v}\cdot\nabla\ln{c_s^2} - \frac{\dd\mathcal{C}}{P},\\
  -\ii\sigma\dd v_r  &= 2\Omega\dd v_\phi + 
  \delta\bm{F}\cdot\hat{\bm{r}},\\
  \ii\sigma\dd v_\phi &= \frac{\kappa^2}{2\Omega}\dd v_r + \frac{\p
    v_\phi}{\p z}\dd v_z,\\
  -\ii\sigma\dd v_z &=  \delta\bm{F}\cdot\hat{\bm{z}},
\end{align} 
with
\begin{align}
  \delta \bm{F} \equiv \frac{\dd\rho}{\rho^2}\nabla P -
  \frac{1}{\rho}\nabla\dd P, 
\end{align}
and $\dd\mathcal{C}$ is the linearized dust-duffusion function (see Eq. \ref{dusty_cooling}).  
Note that for the axisymmetric problem, $\dd\bm{F}$ is purely
meridional.

@@

\section{Linearized pressure forces and its divergence}
Define
\begin{align}
  \bm{F} \equiv - \frac{\nabla P}{\rho}. 
\end{align}
Then its linearized form is 
\begin{align}
  \delta \bm{F} = - \frac{\dd\rho}{\rho}\bm{F} -
  \frac{1}{\rho}\nabla\dd P,
\end{align}
with divergence
\begin{align}
  \nabla\cdot\dd\bm{F} = -
  \nabla\left(\frac{\dd\rho}{\rho}\right)\cdot\bm{F} -
  \frac{\dd\rho}{\rho}\nabla\cdot\bm{F} +
  \nabla\ln{\rho}\cdot\frac{\nabla\dd P}{\rho} -
  \frac{1}{\rho}\nabla^2\dd P. 
\end{align}

\subsection{Radially local approximation}
We assume perturbations have small radial wavelengths compared to the
local disk radius. We take a wave-like dependence $\sim \exp{\ii
  k r}$ for the pertubations with $|kr|\gg 1$ so that curvature terms 
may be neglected. We thus set $\p_r\to \ii k$ when, \emph{and only
  when}, operating on the \emph{primitive} perturbations $\dd P,
\dd\rho$ and $\dd\bm{v}$. %In the following expressions, we only
%apply the radially local approximation at the last step.

\subsection{Explicit expressions}
Introducing
\begin{align}
  W \equiv \frac{\dd\rho}{\rho}, \quad Q \equiv \frac{\dd P}{\rho},
\end{align}
the explicit forms for $\dd\bm{F}$ and $\nabla\cdot\dd\bm{F}$, in the
radially-local approximation, are
\begin{align}
  &\dd F_r = - W F_r - \ii k Q,\\
  &\dd F_z = - W F_z - \left[Q^\prime + Q
    \left(\ln{\rho}\right)^\prime\right]  = s\dd v_z, 
\end{align} 
where $^\prime = \p_z$ and the last equality is the linearized
vertical momentum equation; and 
\begin{align}
  \nabla\cdot\dd\bm{F} &= F_r\left(\p_r\ln{\rho} - \ii k \right)W - F_z
  W^\prime - W \nabla\cdot\bm{F} + \ii kQ \p_r\ln{\rho} +
  \left(\ln{\rho}\right)^\prime\left[Q^\prime + Q
    \left(\ln{\rho}\right)^\prime\right] \notag\\
  & \phantom{=} + k^2 Q - \left\{Q^{\prime\prime} + 2
    Q^\prime\left(\ln{\rho}\right)^\prime
    + Q\left[\left(\ln{\rho}\right)^{\prime\prime}+\left(\ln{\rho}\right)^{\prime
      2}\right]\right\}\notag\\
    &= \left[\left(\p_r\ln{\rho}-\ii k\right)F_r - \nabla\cdot\bm{F} +
      F_z^\prime\right]W + \left(\ii k \p_r\ln{\rho} + k^2\right)Q +
    s\dd v_z^\prime.
\end{align}

\section{Linearized dust diffusion}
We consider small grains in the Epstein regime, with fixed internal
density and size, so that
\begin{align}
  \tstop  =  \frac{K}{\rho c_s}
\end{align}
\citep{price15}, where $K$ is a constant. Then the dust diffusion
function becomes
\begin{align}
  \mathcal{C} \equiv c_s^2\nabla\cdot\left(\epsilon\tstop\nabla
  P\right) = - K c_s^2 \nabla\cdot
  \left(\frac{\epsilon}{c_s}\bm{F}\right) =
  -Kc_s \left(\bm{F}\cdot\nabla\epsilon + \epsilon \nabla\cdot\bm{F}
  - \frac{1}{2}\epsilon\bm{F}\cdot\nabla\ln{c_s^2}\right).  
\end{align}
Linearizing,
\begin{align}
  - \frac{\dd\mathcal{C}}{Kc_s} = \bm{F}\cdot\nabla\dd\epsilon +
  \dd\bm{F}\cdot\nabla\epsilon + \dd\epsilon\nabla\cdot\bm{F} +
  \epsilon\nabla\cdot\dd\bm{F} -
  \frac{1}{2}\nabla\ln{c_s^2}\cdot\left(\bm{F}\dd\epsilon + \epsilon
  \dd\bm{F}\right). 
\end{align}
The linearized dust-fraction $\dd\epsilon$ and its derivatives in the
radially-local approximation are given by
\begin{align}
  \dd\epsilon      &= (1-\epsilon)W - \frac{Q}{c_s^2}, \\
  \p_r\dd\epsilon &= \left[(1-\epsilon)\left(\ii k -
    \p_r\ln{\rho}\right) - \p_r\epsilon\right]W + \left(\p_r\ln{c_s^2}
  + \p_r\ln{\rho} - \ii k \right)\frac{Q}{c_s^2},\\
  \dd\epsilon^\prime &= (1-\epsilon)W^\prime - \epsilon^\prime W -
  \left(\frac{Q}{c_s^2}\right)^\prime. 
\end{align}

>>>>

The linearized equations for axisymmetric disturbances are

\begin{align}
  \ii\sigma\frac{\delta\rho}{\rho} &= \nabla\cdot\dd\bm{v} +
  \dd\bm{v}\cdot\nabla\ln{\rho},\label{lin_mass_full}\\
 \ii\sigma\frac{\delta P}{P} &= \nabla\cdot\dd\bm{v} +
  \dd\bm{v}\cdot\nabla\ln{P} - \dd\bm{v}\cdot\nabla\ln{c_s^2} - \frac{\dd\mathcal{C}}{P}.\label{lin_energy_full}\\
  -\ii\sigma\dd v_r  &= 2\Omega\dd v_\phi + 
  \delta\bm{F}\cdot\hat{\bm{r}},\\
  \ii\sigma\dd v_\phi &= \frac{\kappa^2}{2\Omega}\dd v_r + \frac{\p
    v_\phi}{\p z}\dd v_z,\\
  -\ii\sigma\dd v_z &=  \delta\bm{F}\cdot\hat{\bm{z}}.
\end{align}  
These equations do not assume the radially-local approximation used in
the main text.

@@

\section{Linearized pressure forces and its divergence}\label{lin_press}
Define
\begin{align}
  \bm{F} \equiv - \frac{\nabla P}{\rho}. 
\end{align}
Then its linearized form is 
\begin{align}
  \delta \bm{F} = - \frac{\dd\rho}{\rho}\bm{F} -
  \frac{1}{\rho}\nabla\dd P,
\end{align}
with divergence
\begin{align}
  \nabla\cdot\dd\bm{F} = -
  \nabla\left(\frac{\dd\rho}{\rho}\right)\cdot\bm{F} -
  \frac{\dd\rho}{\rho}\nabla\cdot\bm{F} +
  \nabla\ln{\rho}\cdot\frac{\nabla\dd P}{\rho} -
  \frac{1}{\rho}\nabla^2\dd P. 
\end{align}

%\subsection{Explicit expressions}

The explicit forms for $\dd\bm{F}$ and $\nabla\cdot\dd\bm{F}$, in the
radially-local approximation, are
\begin{align}
  &\dd F_r = - W F_r - \ii k Q,\\
  &\dd F_z = - W F_z - \left[Q^\prime + Q
    \left(\ln{\rho}\right)^\prime\right]  = s\dd v_z, 
\end{align} 
where the last equality is the linearized
vertical momentum equation; and 
\begin{align}
  \nabla\cdot\dd\bm{F} &= F_r\left(\p_r\ln{\rho} - \ii k \right)W - F_z
  W^\prime - W \nabla\cdot\bm{F} + \ii kQ \p_r\ln{\rho} +
  \left(\ln{\rho}\right)^\prime\left[Q^\prime + Q
    \left(\ln{\rho}\right)^\prime\right] \notag\\
  & \phantom{=} + k^2 Q - \left\{Q^{\prime\prime} + 2
    Q^\prime\left(\ln{\rho}\right)^\prime
    + Q\left[\left(\ln{\rho}\right)^{\prime\prime}+\left(\ln{\rho}\right)^{\prime
      2}\right]\right\}\notag\\
    &= \left[\left(\p_r\ln{\rho}-\ii k\right)F_r - \nabla\cdot\bm{F} +
      F_z^\prime\right]W + \left(\ii k \p_r\ln{\rho} + k^2\right)Q +
    s\dd v_z^\prime.
\end{align}

\section{Linearized dust diffusion}\label{lin_dust}
We consider small grains in the Epstein regime, with fixed internal
density and size, so that
\begin{align}
  \tstop  =  \frac{K}{\rho c_s}
\end{align}
\citep{price15}, where $K$ is a constant. Then the dust diffusion
function becomes
\begin{align}
  \mathcal{C} \equiv c_s^2\nabla\cdot\left(\epsilon\tstop\nabla
  P\right) = - K c_s^2 \nabla\cdot
  \left(\frac{\epsilon}{c_s}\bm{F}\right) =
  -Kc_s \left(\bm{F}\cdot\nabla\epsilon + \epsilon \nabla\cdot\bm{F}
  - \frac{1}{2}\epsilon\bm{F}\cdot\nabla\ln{c_s^2}\right).  
\end{align}
Linearizing,
\begin{align}
  - \frac{\dd\mathcal{C}}{Kc_s} = \bm{F}\cdot\nabla\dd\epsilon +
  \dd\bm{F}\cdot\nabla\epsilon + \dd\epsilon\nabla\cdot\bm{F} +
  \epsilon\nabla\cdot\dd\bm{F} -
  \frac{1}{2}\nabla\ln{c_s^2}\cdot\left(\bm{F}\dd\epsilon + \epsilon
  \dd\bm{F}\right). 
\end{align}
The linearized dust-fraction $\dd\epsilon$ and its derivatives in the
radially-local approximation are given by
\begin{align}
  \dd\epsilon      &= (1-\epsilon)W - \frac{Q}{c_s^2}, \\
  \p_r\dd\epsilon &= \left[(1-\epsilon)\left(\ii k -
    \p_r\ln{\rho}\right) - \p_r\epsilon\right]W + \left(\p_r\ln{c_s^2}
  + \p_r\ln{\rho} - \ii k \right)\frac{Q}{c_s^2},\\
  \dd\epsilon^\prime &= (1-\epsilon)W^\prime - \epsilon^\prime W -
  \left(\frac{Q}{c_s^2}\right)^\prime. 
\end{align}

### 2016-05-06 model.tex
updated notes

\begin{align}
\frac{DP}{Dt} = - P \nabla\cdot\bm{v} + P\bm{v}\cdot\nabla\ln{c_s^2} +
c_s^2 \nabla\cdot\left[\tstop\left(1 - \frac{P}{c_s^2\rho}\right)\nabla
  P\right]. \label{eff_energy}
\end{align}

This is the energy equation for an ideal gas of adiabatic index
$\gamma=1$, i.e. isothermal evolution, but now with additional source 
terms due to the imposed temperature profile, as well as dust-gas
friction.

@@

\subsection{Locally isothermal gas perfectly coupled to dust} 
When $c_s(r,z)$ is non-uniform but $\tstop=0$, simple stability
criteria are not available. However, we may obtain an integral
relation for linear, axisymmetric waves with frequency $\sigma$ (see
Appendix \ref{var_prin}) in the form:  
\begin{align}
&  \sigma^2\int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV \notag\\
&= \int\left[ \rho
  |\dd v_r|^2A + \rho  \dd v_z \dd v_r^* B + \rho \dd v_z^*\dd v_r C +
  \rho |\dd v_z|^2 D\phantom{\frac{1}{1}}\right. \notag\\
&\phantom{===}  \left. + \frac{1}{P}\left|\nabla\cdot\left(P\dd
  \bm{v}\right)\right|^2\right]dV - \int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV.
\end{align}
The real coefficients $A,B,C,D$ can be read off
Eq. \ref{integral_ex}. Note that $B=C$ from the assume equilibrium
structure (Eq. \ref{vshear}).

@@

\subsection{Strictly isothermal gas well-coupled to dust} 
Now consider $c_s^2=\mathrm{const.}$ but $\tstop\neq0$. In this case
the effective energy equation is 
\begin{align}\label{dusty_cooling}
  \frac{DP}{Dt} &= - P \nabla \cdot \bm{v} + \mathcal{C},\\
  \mathcal{C} & = \nabla\cdot\left[\tstop\left(c_s^2 -
    \frac{P}{\rho}\right)\nabla P\right].           
\end{align} 
The function $\mathcal{C}$ can be considered as a cooling term. By
analogy with enabling VSI in an adiabatic gas through cooling
\citep{nelson13,lin15}, we may expect a `dusty VSI' to develop 
if this term is large enough (physically meaning diffusion of the
dust-fraction is strong).

>>>>

\begin{align}
\frac{DP}{Dt} &= - P \nabla\cdot\bm{v} + P\bm{v}\cdot\nabla\ln{c_s^2}
                + \mathcal{C},  \label{eff_energy} \\
\mathcal{C}&\equiv c_s^2 \nabla\cdot\left[\tstop\left(1 - \frac{P}{c_s^2\rho}\right)\nabla
  P\right]
\end{align}

Eq. \ref{eff_energy} is the energy equation for an ideal gas of adiabatic index
$\gamma=1$, i.e. isothermal evolution, but now with additional source 
terms due to the imposed temperature profile, as well as dust-gas
friction.

@@

\subsection{Locally isothermal gas perfectly coupled to dust} 
When $c_s(r,z)$ is non-uniform but $\tstop=0$, simple stability
criteria are not available. However, we may obtain an integral
relation for linear, axisymmetric waves with frequency $\sigma$ (see
Appendix \ref{var_prin}, where we also account for finite $\tstop$) in
the form:   
\begin{align}
&  \sigma^2\int\rho\left(|\dd v_r|^2 + |\dd v_z|^2\right)dV \notag\\
&= \int\left[ \rho
  |\dd v_r|^2A + \rho  \dd v_z \dd v_r^* B + \rho \dd v_z^*\dd v_r C +
  \rho |\dd v_z|^2 D\phantom{\frac{1}{1}}\right. \notag\\
&\phantom{===}  \left. + \frac{1}{P}\left|\nabla\cdot\left(P\dd
  \bm{v}\right)\right|^2\right]dV - \int P
  \left(\nabla\cdot\dd\bm{v}^*\right)\left(\dd\bm{v}\cdot\nabla\ln{c_s^2}\right)dV.
\end{align}
The real coefficients $A,B,C,D$ can be read off
Eq. \ref{integral_ex}. Note that $B=C$ from the assume equilibrium
structure (Eq. \ref{vshear}).

@@

\subsection{Strictly isothermal gas well-coupled to dust} 
Now consider $c_s^2=\mathrm{const.}$ but $\tstop\neq0$. In this case
the effective energy equation is 
\begin{align}\label{dusty_cooling}
  \frac{DP}{Dt} &= - P \nabla \cdot \bm{v} + \mathcal{C}
\end{align} 
The function $\mathcal{C}$ can be considered as a cooling term. By
analogy with enabling VSI in an adiabatic gas through cooling
\citep{nelson13,lin15}, we may expect a `dusty VSI' to develop 
if this term is large enough (physically meaning diffusion of the
dust-fraction is strong).
